---
title: "Standardization"
author: "Brooke Wolford"
date: "2025-02-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(RNOmni)
set.seed(1234)

```

# Simulate data
```{r}
#make a messy distribution by mixing normal and uniform
x<-rnorm(100,mean=20,sd=100)
y<-runif(100)*100
a<-c(x,y)

hist(a)
mean(a)
var(a)
sd(a)
```

# Standardize
Scale does nothing to make the data normal. Makes mean of 0 and variance of 1. Using the default settings in R it scales each element by subtracting the mean and dividing by the SD.  
```{r}
###standardize (substracts mean and divides by sd)
scaled<-scale(a)
hist(scaled)
mean(scaled)
var(scaled)
sd(scaled)
```

# Normalization (Min-max) 
A range based version also exists. This re-scales the data to a range between 0 and 1. This technique preserves the relative order and distance of the data points, but it also reduces the variance and magnifies the effect of outliers
```{r echo=FALSE}
minMax <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}
mm<-minMax(a)
hist(mm)
mean(mm)
var(mm)
sd(mm)
```

# Rank based inverse normal transformation
Essentially this first ranks the data, then converts them to z-scores using an inverse-normal transformation. This makes the data normally distributed as well as having a mean of 1 and SD of 1. A non-rank version also exists. Read more about INT [here](https://yingji15.github.io/2019-08-17-inverse-normal-transformation/) and [here](https://pmc.ncbi.nlm.nih.gov/articles/PMC8643141/).  
```{r echo=FALSE}
### makes mean of 0 and variance of 1 while also making the distribution normal 
ranked<-RankNorm(a)
hist(ranked)
mean(ranked)
var(ranked)
sd(ranked)
```

